Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms

### 2014-03-26 CGP Journal/main.tex
First full draft of response to reviewers
Changes to emphasize "boolean" nature of experiments.

Our results show that most CGP techniques evolve genomes that
are highly inactive, very redundant, and full of seemingly useless constants.
On some tested problems we found less than 1\% of the genome was actually required to
encode the evolved solution.
Furthermore, traditional CGP ordering results in large portions of the genome
that are never used by any ancestor of the evolved solution.
\emph{Reorder} and \emph{DAG} allow evolution to utilize the entire genome.
More generally, our
results suggest that \emph{Skip-Reorder} and \emph{Single-Reorder}
are most likely to solve hard problems using the least number of evaluations and
the least amount of time while better avoiding degenerate
behavior.

@@

In this work we set out to develop a deeper understanding of what makes Cartesian
Genetic Programming (CGP) an effective evolutionary optimizer.
Previous studies of CGP's evolutionary mechanisms have
focused on bloat~\cite{miller:2001:bloat},
neutrality~\cite{vassilev:2000:neutrality}, and structural bias~\cite{payne:2009:bias}.
We shall focus on the interaction between mutation and genome ordering.
%NEW
While these features of CGP have trivial definitions, we will argue, using
experimentation and analysis of alternatives, that they have a great deal
of impact on search success.
%/NEW

@@

CGP uses very simple evolutionary mechanisms.  The most common evolutionary strategy
is $\mu + \lambda$ where $\mu \leftarrow 1$ and $\lambda \leftarrow 4$.
This means during each generation a single parent produces four offspring using
mutation.  The best offspring then competes with the parent,\footnote{Ties are broken
randomly.} with the offspring
replacing the parent if it is no less fit.  This replacement strategy encourages
neutral drift.
%NEW
In CGP mutation is done in a myriad of nearly synonymous ways.  Here we mutate
%/NEW
each gene of each node at a set probability, where mutation
involves changing a gene randomly to some different valid value.  For example,
if a function gene is chosen for mutation, its new value is randomly chosen from
all possible functions, excluding the gene's current value.
%NEW
This version of mutation was chosen to allow precise prediction of changes
in gene value necessarily in later sections.
%/NEW
Combined,
this form of CGP can be viewed as a stochastic hill climber with neutrality.
For a more in depth description of CGP, see~\cite{miller:2011:chapter2}.

@@

%NEW
This technique incurs a minor increase in runtime, as each active gene in the
offspring must be compared against the parent's gene at the same locus.
This operation is at most linear in genome size.  Yet
on problems where evaluation is even somewhat expensive, the evaluations prevented
by performing this check will result in a net reduction in runtime.
%/NEW

@@

%NEW
Similar to \emph{Skip}, \emph{Accumulate} incurs a minor computation cost in
order to prevent wasting evaluations.  Generating each $F_i$ requires at most
$O(AN)$, where $A$ is the arity of the nodes and $N$ is the number of nodes.
While \emph{Accumulate} may produce many $F_i$ in the process of creating
a single offspring, conceptually \emph{Skip} using the same mutation rate
will create a similar number of total individuals when performing the same
number of evaluations.  As such, we suggest \emph{Accumulate}
will take no more than a constant factor more runtime to produce the same number
of evaluable individuals as \emph{Skip}.  As before, this increase is likely
eclipsed when solving problems with sufficiently time consuming evaluation functions.
%/NEW

@@

%NEW
Conceptually this method could be extended to mutate some number of active genes
before terminating.  Doing so would reintroduce a parameter to govern how many
active genes to mutate.  Our choice of a single mutation allows for the minimal
amount of change to ensure the individual should be evaluated, while mutations
made to inactive genes still give CGP sufficient power in each mutation to escape
local optimums.
%/NEW

@@

To rectify these issues and examine CGP without ordering bias, \cite{goldman:2013:ordering}
created \emph{Reorder} and \emph{DAG} as methods for genome ordering.
%NEW
For clarity, we shall refer to CGP's historical ordering method of enforcing
only forward connections as \emph{Normal}.
%/NEW

%NEW
It is important to note that Equation~\ref{eq:prob} does not apply to all variants of CGP.
For instance, the early versions of CGP~\cite{miller:2011:chapter2} which utilize a
levels back parameter $L$ will have
a flat active probability for all nodes at least $L$ nodes from the start and $L$ nodes from the end
of the genome.  Yet this does not remove the issue of artificially limiting connections,
and in fact adds many new limitations.  More recently,~\cite{cai:2005:ircgp} proposed
a method which reassembled the genome using semantic cues, which circumvents the
forward only limitation in \emph{Normal}.
%/NEW
\subsection{Reorder}
\label{sec:reorder}
The \emph{Reorder} method shuffles an existing genome's node ordering without
changing node behavior.  This is possible because for a given node, there can
be a large number of other nodes which have no required ordering,
as they neither take input (directly or indirectly) from the node nor provide their
output (directly or indirectly) to the node.  In general, this process works by assigning nodes new
locations in the genome at random once all of the nodes they take input from have been assigned
locations earlier in the genome.

@@

The price \emph{Reorder} pays for this potential improvement is a larger mutational search
space.  Though a shuffled individual is phenotypically unchanged,
its mutational adjacency can be significantly different.  As such the variety of offspring
an individual can create is increased when using \emph{Reorder}.
%/NEW
Therefore creating a specific offspring that improves upon the parent is
less likely, in that the number of alternatives is increased.  Yet many
of these additional alternatives may also be improvements upon the parent.  This may
increase the likelihood of producing an offspring that improves
upon the parent, allowing \emph{Reorder} to overcome this drawback.
%/NEW

@@

As in \emph{Normal} and \emph{Reorder}, when a connection gene is chosen for mutation, its value is chosen randomly.
However, the options to choose from are only limited to those connections which do not
create cycles.
To achieve this, we incrementally determine
%NEW No change was made to mutating node as that is in fact what we wanted
%/NEW
which nodes transitively depend on the mutating node and which do not, with
information stored in a $known$ map data structure.  This data structure is initialized with
the information that the node is dependent on itself and that all input locations
are not dependent on the mutating node.  Nodes are then
tested in a random order for dependence using the \Call{IsDependent}{} algorithm given in Figure~\ref{fig:dag}.
The first node returned at the top level that is not dependent on the mutating node is then
used as the new connection gene value.  As a result, mutating a connection gene
cannot introduce a cycle.

@@

To choose which parameter configuration to use with each algorithm combination
for each problem, we used an iterative process of performing runs, comparing medians, removing configurations,
and repeating.  For each algorithm combination and problem, we first ran all
potential parameter configurations 5 times (320 runs if not using \emph{Single}, 40 otherwise).
The parameter configurations are then sorted based on their median.  The best half are then
each run 4 more times, with the best half of that group chosen.  In this way, 32 configurations
are run 5 times, 16 are run 9 times, 8 are run 13 times, 4 are run 17 times, 2 are run 21 times,
and 2 are run 25 times.  For algorithm combinations using \emph{Single}, all 8 parameter configurations were run 17 times
before any were removed.  This iterative reduction allows us to focus tuning on
those parameter configurations most likely to be effective, with more information
gathered before choosing between high quality configurations.
%NEW
We chose to run each initially 5 times to reduce the chance of
eliminating a configuration due to outliers.  We incrementally added 4
to provide enough information to significantly update estimates,
while keeping the number of runs odd, allowing the median to be a
representative run and not an interpolation.  As a final consideration
the number of experiments were set to make the total number of runs feasible.
%/NEW
In total this means
15,696 complete runs are required to set the parameter configuration for all
algorithm configurations for all problems.

@@

The results from running each tuned parameter configuration for each algorithm
combination on each problem are summarized by Table~\ref{tab:results}.
Rows for each problem are ordered by that algorithm combination's median evaluations to
success (MES), with the best algorithm combination appearing at the top of each section.
The highlighted rows mark \emph{Skip-Normal}, the control algorithm combination
representing traditional CGP performance.  All configurations for each problem
were then tested using
%NEW
Kruskal-Wallis\footnote{Kruskal-Wallis, a non-parametric equivalent of ANOVA, determines
if samples were drawn from the same distribution.  If this test is significant, at
least one of the samples differs from the others.}
%/NEW
to determine if any configuration was
statistically different from any other.  All problems were determined to be highly significant.
As such, each algorithm combination was then compared with the control
to determine statistical significance using pairwise
%NEW
Mann-Whitney~U\footnote{Mann-Whitney~U, a non-parametric equivalent of Student's t-test,
determines if two populations were drawn from the same distribution.  If this test is significant,
the populations are not equal.}
%/NEW
tests. The
resulting p-value for each test is reported in the table.

@@

The difference between using \emph{Skip} and \emph{Accumulate} remains
negligible~\cite{goldman:2013:cgpwaste}, even in combination with different ordering
methods.
%NEW
This supports the idea that the two are actually achieving synonymous behavior
through different means.
%/NEW
As \emph{Accumulate} is far more complex in terms of algorithm function
than \emph{Skip}, we would suggest \emph{Skip} be used over \emph{Accumulate} for
%NEW Slight change here, "real world" became "future"
any future applications.

As a final check that our results do not somehow unfairly handicap traditional CGP,
we compared our median evaluations to success with previous publications.  The results
reported for traditional CGP in~\cite{walker:2008:cgpmodules}
for 3-bit Parity required 6 times as many evaluations as \emph{Skip-Normal}.
On the 3-bit Multiplier problem using a slightly different operator set,
the same source had traditional CGP requiring 4 million evaluations to success,
as opposed to our \emph{Skip-Normal} requiring 600 thousand, also a factor of 6 improvement.
%NEW
To determine the effect the difference in function set had on results, we ran
\emph{Skip-Normal} and \emph{Skip-Reorder} using their functions but our parameters.
The former improved to 453,684 while the latter increased to 303,657, maintaining
a statistically significant difference.  These
results make sense as the operator set in~\cite{walker:2008:cgpmodules} is able
to more compactly represent the solution than our set.  This puts less strain
on \emph{Normal} to deal with complex solutions, while \emph{Reorder}'s extremely
large genome size is likely made unnecessary.  Note that as we did not tune
using their operator set, our results should be considered a lower bound on quality for
using \emph{Skip-Normal} and \emph{Skip-Reorder} to solve 3-bit Multiplier
with their operators.
%/NEW

>>>>

%TODO mention this
Our results on boolean problems show CGP evolves genomes that
are highly inactive, very redundant, and full of seemingly useless constants.
On some tested problems we found less than 1\% of the genome was actually required to
encode the evolved solution.
Furthermore, traditional CGP ordering results in large portions of the genome
that are never used by any ancestor of the evolved solution.
\emph{Reorder} and \emph{DAG} allow evolution to utilize the entire genome.
More generally, our
results suggest that \emph{Skip-Reorder} and \emph{Single-Reorder}
are most likely to solve hard problems using the least number of evaluations and
the least amount of time while better avoiding degenerate
behavior.

@@

In this work we set out to develop a deeper understanding of what makes Cartesian
Genetic Programming (CGP) an effective evolutionary optimizer.
Previous studies of CGP's evolutionary mechanisms have
focused on bloat~\cite{miller:2001:bloat},
neutrality~\cite{vassilev:2000:neutrality}, and structural bias~\cite{payne:2009:bias}.
We shall focus on the interaction between mutation and genome ordering, both
%TODO final change
generally and how they specifically apply to the boolean problem domain.
%Done-NEW
While these features of CGP have trivial definitions, we will argue, using
experimentation and analysis of alternatives, that they have a great deal
of impact on search success.
%/NEW

@@

CGP uses very simple evolutionary mechanisms.  The most common evolutionary strategy
is $\mu + \lambda$ where $\mu \leftarrow 1$ and $\lambda \leftarrow 4$.
This means during each generation a single parent produces four offspring using
mutation.  The best offspring then competes with the parent,\footnote{Ties are broken
randomly.} with the offspring
replacing the parent if it is no less fit.  This replacement strategy encourages
neutral drift.
%Done-NEW
In CGP mutation is done in a myriad of nearly synonymous ways.  Here we mutate
%/NEW
each gene of each node at a set probability, where mutation
involves changing a gene randomly to some different valid value.  For example,
if a function gene is chosen for mutation, its new value is randomly chosen from
all possible functions, excluding the gene's current value.
%Done-NEW
This version of mutation was chosen to allow precise prediction of changes
in gene value necessarily in later sections.
%/NEW
Combined,
this form of CGP can be viewed as a stochastic hill climber with neutrality.
For a more in depth description of CGP, see~\cite{miller:2011:chapter2}.

@@

%Done-NEW
This technique incurs a minor increase in runtime, as each active gene in the
offspring must be compared against the parent's gene at the same locus.
This operation is at most linear in genome size.  Yet
on problems where evaluation is even somewhat expensive, the evaluations prevented
by performing this check will result in a net reduction in runtime.
%/NEW

@@

%Done-NEW
Similar to \emph{Skip}, \emph{Accumulate} incurs a minor computation cost in
order to prevent wasting evaluations.  Generating each $F_i$ requires at most
$O(AN)$, where $A$ is the arity of the nodes and $N$ is the number of nodes.
While \emph{Accumulate} may produce many $F_i$ in the process of creating
a single offspring, conceptually \emph{Skip} using the same mutation rate
will create a similar number of total individuals when performing the same
number of evaluations.  As such, we suggest \emph{Accumulate}
will take no more than a constant factor more runtime to produce the same number
of evaluable individuals as \emph{Skip}.  As before, this increase is likely
eclipsed when solving problems with sufficiently time consuming evaluation functions.
%/NEW

@@

%Done-NEW
Conceptually this method could be extended to mutate some number of active genes
before terminating.  Doing so would reintroduce a parameter to govern how many
active genes to mutate.  Our choice of a single mutation allows for the minimal
amount of change to ensure the individual should be evaluated, while mutations
made to inactive genes still give CGP sufficient power in each mutation to escape
local optimums.
%/NEW

@@

To rectify these issues and examine CGP without ordering bias, \cite{goldman:2013:ordering}
created \emph{Reorder} and \emph{DAG} as methods for genome ordering.
%Done-NEW
For clarity, we shall refer to CGP's historical ordering method of enforcing
only forward connections as \emph{Normal}.
%/NEW

%DONE-NEW
It is important to note that Equation~\ref{eq:prob} does not apply to all variants of CGP.
For instance, the early versions of CGP~\cite{miller:2011:chapter2} which utilize a
levels back parameter $L$ will have
a flat active probability for all nodes at least $L$ nodes from the start and $L$ nodes from the end
of the genome.  Yet this does not remove the issue of artificially limiting connections,
and in fact adds many new limitations.  More recently,~\cite{cai:2005:ircgp} proposed
a method which reassembled the genome using semantic cues, which circumvents the
forward only limitation in \emph{Normal}.
%/NEW
\subsection{Reorder}
\label{sec:reorder}
The \emph{Reorder} method shuffles an existing genome's node ordering without
changing node behavior.  This is possible because for a given node, there can
be a large number of other nodes which have no required ordering,
as they neither take input (directly or indirectly) from the node nor provide their
output (directly or indirectly) to the node.  In general, this process works by assigning nodes new
locations in the genome at random once all of the nodes they take input from have been assigned
locations earlier in the genome.

@@

The price \emph{Reorder} pays for this potential improvement is a larger mutational search
space.  Though a shuffled individual is phenotypically unchanged,
its mutational adjacency can be significantly different.  As such the variety of offspring
an individual can create is increased when using \emph{Reorder}.
%Done-NEW
Therefore creating a specific offspring that improves upon the parent is
less likely, in that the number of alternatives is increased.  Yet many
of these additional alternatives may also be improvements upon the parent.  This may
increase the likelihood of producing an offspring that improves
upon the parent, allowing \emph{Reorder} to overcome this drawback.
%/NEW

@@

As in \emph{Normal} and \emph{Reorder}, when a connection gene is chosen for mutation, its value is chosen randomly.
However, the options to choose from are only limited to those connections which do not
create cycles.
To achieve this, we incrementally determine
%Done-NEW No change was made to mutating node as that is in fact what we wanted
%/NEW
which nodes transitively depend on the mutating node and which do not, with
information stored in a $known$ map data structure.  This data structure is initialized with
the information that the node is dependent on itself and that all input locations
are not dependent on the mutating node.  Nodes are then
tested in a random order for dependence using the \Call{IsDependent}{} algorithm given in Figure~\ref{fig:dag}.
The first node returned at the top level that is not dependent on the mutating node is then
used as the new connection gene value.  As a result, mutating a connection gene
cannot introduce a cycle.

@@

To choose which parameter configuration to use with each algorithm combination
for each problem, we used an iterative process of performing runs, comparing medians, removing configurations,
and repeating.  For each algorithm combination and problem, we first ran all
potential parameter configurations 5 times (320 runs if not using \emph{Single}, 40 otherwise).
The parameter configurations are then sorted based on their median.  The best half are then
each run 4 more times, with the best half of that group chosen.  In this way, 32 configurations
are run 5 times, 16 are run 9 times, 8 are run 13 times, 4 are run 17 times, 2 are run 21 times,
and 2 are run 25 times.  For algorithm combinations using \emph{Single}, all 8 parameter configurations were run 17 times
before any were removed.  This iterative reduction allows us to focus tuning on
those parameter configurations most likely to be effective, with more information
gathered before choosing between high quality configurations.
%Done-NEW
We chose to run each initially 5 times to reduce the chance of
eliminating a configuration due to outliers.  We incrementally added 4
to provide enough information to significantly update estimates,
while keeping the number of runs odd, allowing the median to be a
representative run and not an interpolation.  As a final consideration
the number of experiments were set to make the total number of runs feasible.
%/NEW
In total this means
15,696 complete runs are required to set the parameter configuration for all
algorithm configurations for all problems.

@@

The results from running each tuned parameter configuration for each algorithm
combination on each problem are summarized by Table~\ref{tab:results}.
Rows for each problem are ordered by that algorithm combination's median evaluations to
success (MES), with the best algorithm combination appearing at the top of each section.
The highlighted rows mark \emph{Skip-Normal}, the control algorithm combination
representing traditional CGP performance.  All configurations for each problem
were then tested using
%Done-NEW
Kruskal-Wallis\footnote{Kruskal-Wallis, a non-parametric equivalent of ANOVA, determines
if samples were drawn from the same distribution.  If this test is significant, at
least one of the samples differs from the others.}
%/NEW
to determine if any configuration was
statistically different from any other.  All problems were determined to be highly significant.
As such, each algorithm combination was then compared with the control
to determine statistical significance using pairwise
%Done-NEW
Mann-Whitney~U\footnote{Mann-Whitney~U, a non-parametric equivalent of Student's t-test,
determines if two populations were drawn from the same distribution.  If this test is significant,
the populations are not equal.}
%/NEW
tests. The
resulting p-value for each test is reported in the table.

@@

The difference between using \emph{Skip} and \emph{Accumulate} remains
negligible~\cite{goldman:2013:cgpwaste}, even in combination with different ordering
methods.
%Done-NEW
This supports the idea that the two are actually achieving synonymous behavior
through different means.
%/NEW
As \emph{Accumulate} is far more complex in terms of algorithm function
than \emph{Skip}, we would suggest \emph{Skip} be used over \emph{Accumulate} for
%Done-NEW Slight change here, "real world" became "future"
any future applications.

As a final check that our results do not somehow unfairly handicap traditional CGP,
we compared our median evaluations to success with previous publications.  The results
reported for traditional CGP in~\cite{walker:2008:cgpmodules}
for 3-bit Parity required 6 times as many evaluations as \emph{Skip-Normal}.
On the 3-bit Multiplier problem using a slightly different operator set,
the same source had traditional CGP requiring 4 million evaluations to success,
as opposed to our \emph{Skip-Normal} requiring 600 thousand, also a factor of 6 improvement.
%DONE-NEW
To determine the effect the difference in function set had on results, we ran
\emph{Skip-Normal} and \emph{Skip-Reorder} using their functions but our parameters.
The former improved to 453,684 while the latter increased to 303,657, maintaining
a statistically significant difference.  These
results make sense as the operator set in~\cite{walker:2008:cgpmodules} is able
to more compactly represent the solution than our set.  This puts less strain
on \emph{Normal} to deal with complex solutions, while \emph{Reorder}'s extremely
large genome size is likely made unnecessary.  Note that as we did not tune
using their operator set, our results should be considered a lower bound on quality for
using \emph{Skip-Normal} and \emph{Skip-Reorder} to solve 3-bit Multiplier
with their operators.
%/NEW
